---
title: "DATA 606 Data Project Proposal"
author: "Eric Lehmphul"
date: "10/31/2021"
output: pdf_document
---

```{r include = FALSE}
knitr::opts_chunk$set(fig.align="center", fig.width = 6)
```

# Data Preparation

Retrieved dataset from Kaggle: https://www.kaggle.com/sulianova/cardiovascular-disease-dataset. It is a dataset relating to cardiovascular disease and relative variables of interest.

```{r}
library(tidyverse)

# load data

cardio.data <- read.csv("https://raw.githubusercontent.com/SaneSky109/DATA606/main/Data_Project/Data/cardio_train.csv", sep = ";")
```

```{r}
# remove unecessary column: id
cardio.data <- cardio.data[,-1]
```

```{r}
# create factors
cardio.data$cardio <- factor(cardio.data$cardio)
cardio.data$gender <- factor(cardio.data$gender)
cardio.data$cholesterol <- factor(cardio.data$cholesterol)
cardio.data$gluc <- factor(cardio.data$gluc)
cardio.data$smoke <- factor(cardio.data$smoke)
cardio.data$alco <- factor(cardio.data$alco)
cardio.data$active <- factor(cardio.data$active)
```

```{r}
# rename factor levels

levels(cardio.data$cardio) <- c("No", "Yes")
levels(cardio.data$gender) <- c("Female", "Male")
levels(cardio.data$cholesterol) <- c("Normal", "Above_Normal", "Well_Above_Normal")
levels(cardio.data$gluc) <- c("Normal", "Above_Normal", "Well_Above_Normal")
levels(cardio.data$smoke) <- c("No", "Yes")
levels(cardio.data$alco) <- c("No", "Yes")
levels(cardio.data$active) <- c("No", "Yes")
```

```{r}
# transform age since it is in days

cardio.data$age <- cardio.data$age/365

# remove outliers of ap_hi

# I am assuming the that these measures are errors and 
# I am just dropping them due to problems it will cause with modeling 
# Highest pressure recorded in an individual was 370/360.(https://pubmed.ncbi.nlm.nih.gov/7741618/)

summary(cardio.data$ap_hi)

cardio.data <- cardio.data[cardio.data$ap_hi <= 370,]
cardio.data <- cardio.data[cardio.data$ap_hi > 0,]

summary(cardio.data$ap_hi)


# remove outliers of ap_lo


summary(cardio.data$ap_lo)

cardio.data <- cardio.data[cardio.data$ap_lo <= 360,]
cardio.data <- cardio.data[cardio.data$ap_lo > 0,]

summary(cardio.data$ap_lo)
```


```{r}
glimpse(cardio.data)
```


# Research question 

My research question is: **Do gender, age, body weight, body height, blood pressure, cholesterol, glucose levels, smoking, drinking alcohol and activity level of an individual significantly influence the likelihood of contracting cardiovascular disease?**

I aim to determine what variables are the most important determining factors to cardiovascular disease given the data presented in the dataset.


# Cases 

The cases are the number of people who participate in the medical examination. There were a total of 70,000 cases in the original data file. After data pre-processing, the number of cases is 68,985. This change is due to the removal of rows that seemed to be errors such as extremely high and low blood pressure (-1,000 or 15,000). 

```{r}
nrow(cardio.data)
```


# Data collection 

The data was collected from medical information given by patient and examination results. "All of the dataset values were collected at the moment of medical examination." (https://www.kaggle.com/sulianova/cardiovascular-disease-dataset)

The data was downloaded from Kaggle (https://www.kaggle.com/sulianova/cardiovascular-disease-dataset) and then I uploaded it to Github to be used to import the data into R.

# Type of study 

This is an observational study since the analysis is on events that have already occurred. 

# Data Source 

The link to where I retrieved the data is: https://www.kaggle.com/sulianova/cardiovascular-disease-dataset

# Dependent Variable

The response variable is `cardio`. This is a qualitative variable since it is a categorical binary variable. `cardio` is an indicator variable that indicates whether or not someone has cardiovascular disease.

# Independent Variable

There are multiple variables that I am considering for analysis. The list contains a group of both quantitative and qualitative variables:

* `age` (quantitative): Age of patient in years
* `gender` (qualitative): Gender of patient
* `height` (quantitative): Height of patient in cm
* `weight` (quantitative): Weight of patient in kg
* `ap_hi` (quantitative): Systolic blood pressure
* `ap_lo` (quantitative): Diastolic blood pressure
* `cholesterol` (qualitative): Cholesterol level of patient
* `smoke` (qualitative): Binary variable to determine if a patient smokes
* `alco` (qualitative): Binary variable to determine if a patient drinks alcohol
* `gluc` (qualitative): Glucose level of patient
* `active` (qualitative): Yes/No if patient is physically active

# Relevant Summary Statistics 

## Summary Statistics

```{r}
summary(cardio.data)
```

--------------------------------------------------------------------------------

## Visualizations

### Cardiovascular Disease Outcome by Age

```{r}
ggplot(cardio.data, aes(x=cardio, y=age)) +
  geom_boxplot() +
  ggtitle("Age by Cardiovascular Disease Level")

plot(cardio.data$cardio ~ cardio.data$age, xlab = "Age",ylab = "Cardio", main = "Cardiovascular Outcome vs. Age")
```


--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Gender

```{r}
plot(cardio.data$cardio ~ cardio.data$gender, xlab = "Gender",ylab = "Cardio", main = "Cardiovascular Outcome vs. Gender")

cardio.data %>%
  group_by(gender) %>%
  count(cardio)
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Height

```{r}
ggplot(cardio.data, aes(x=cardio, y=height)) +
  geom_boxplot() +
  ggtitle("Body Height (cm) by Cardiovascular Disease Level")

plot(cardio.data$cardio ~ cardio.data$height, xlab = "Height",ylab = "Cardio", main = "Cardiovascular Outcome vs. Height")
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Weight

```{r}
ggplot(cardio.data, aes(x=cardio, y=weight)) +
  geom_boxplot() +
  ggtitle("Body Weight (kg) by Cardiovascular Disease Level")

plot(cardio.data$cardio ~ cardio.data$weight, xlab = "Weight",ylab = "Cardio", main = "Cardiovascular Outcome vs. Weight")
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Systolic blood pressure

```{r}
ggplot(cardio.data, aes(x=cardio, y=ap_hi)) +
  geom_boxplot() +
  ggtitle("Systolic Blood Pressure by Cardiovascular Disease Level")

plot(cardio.data$cardio ~ cardio.data$ap_hi, xlab = "Systolic Blood Pressure",ylab = "Cardio", main = "Cardiovascular Outcome vs. Systolic Blood Pressure")
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Diastolic blood pressure

```{r}
ggplot(cardio.data, aes(x=cardio, y=ap_lo)) +
  geom_boxplot() +
  ggtitle("Diastolic Blood Pressure by Cardiovascular Disease Level")

plot(cardio.data$cardio ~ cardio.data$ap_hi, xlab = "Diastolic Blood Pressure",ylab = "Cardio", main = "Cardiovascular Outcome vs. Diastolic Blood Pressure")
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Cholesterol Level

```{r}
plot(cardio.data$cardio ~ cardio.data$cholesterol, xlab = "Cholesterol Level",ylab = "Cardio", main = "Cardiovascular Outcome vs. Cholesterol Level")
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Glucose Level

```{r}
plot(cardio.data$cardio ~ cardio.data$gluc, xlab = "Glucose Level",ylab = "Cardio", main = "Cardiovascular Outcome vs. Glucose Level")
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Smoking Level

```{r}
plot(cardio.data$cardio ~ cardio.data$smoke, xlab = "Smoking",ylab = "Cardio", main = "Cardiovascular Outcome vs. Smoking")
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Alcohol Level

```{r}
plot(cardio.data$cardio ~ cardio.data$alco, xlab = "Alcohol",ylab = "Cardio", main = "Cardiovascular Outcome vs. Alcohol Level")
```

--------------------------------------------------------------------------------

\clearpage

### Cardiovascular Disease Outcome by Activity Level

```{r}
plot(cardio.data$cardio ~ cardio.data$active, xlab = "Active",ylab = "Cardio", main = "Cardiovascular Outcome vs. Activity Level")
```

--------------------------------------------------------------------------------


