---
title: "DATA 606 Data Project Proposal"
author: "Eric Lehmphul"
date: "10/31/2021"
output:
  html_document:
    theme: lumen
    df_print: paged
  pdf_document: default
---

```{r include = FALSE}
knitr::opts_chunk$set(fig.align="center", fig.width = 6)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Data Preparation

Retrieved dataset from Kaggle: https://www.kaggle.com/sulianova/cardiovascular-disease-dataset. It is a dataset relating to cardiovascular disease and relative variables of interest.

```{r}
library(tidyverse)
library(GGally)

# load data
url <- "https://raw.githubusercontent.com/SaneSky109/DATA606/main/Data_Project/Data/cardio_train.csv"
cardio.data <- read.csv(url, sep = ";")
```

```{r}
# remove unecessary column: id
cardio.data <- cardio.data[,-1]
```

```{r}
# create factors
cardio.data$cardio <- factor(cardio.data$cardio)
cardio.data$gender <- factor(cardio.data$gender)
cardio.data$cholesterol <- factor(cardio.data$cholesterol)
cardio.data$gluc <- factor(cardio.data$gluc)
cardio.data$smoke <- factor(cardio.data$smoke)
cardio.data$alco <- factor(cardio.data$alco)
cardio.data$active <- factor(cardio.data$active)
```

```{r}
# rename factor levels

levels(cardio.data$cardio) <- c("No", "Yes")
levels(cardio.data$gender) <- c("Female", "Male")
levels(cardio.data$cholesterol) <- c("Norm", "Higher", "Highest")
levels(cardio.data$gluc) <- c("Norm", "Higher", "Highest")
levels(cardio.data$smoke) <- c("No", "Yes")
levels(cardio.data$alco) <- c("No", "Yes")
levels(cardio.data$active) <- c("No", "Yes")
```

```{r}
# transform age since it is in days

cardio.data$age <- cardio.data$age/365

# remove outliers of ap_hi

# I am assuming the that these measures are errors and 
# I am just dropping them due to problems it will cause with modeling 
# Highest pressure recorded in an individual was 370/360.(https://pubmed.ncbi.nlm.nih.gov/7741618/)

summary(cardio.data$ap_hi)

cardio.data <- cardio.data[cardio.data$ap_hi <= 370,]
cardio.data <- cardio.data[cardio.data$ap_hi > 0,]

summary(cardio.data$ap_hi)


# remove outliers of ap_lo


summary(cardio.data$ap_lo)

cardio.data <- cardio.data[cardio.data$ap_lo <= 360,]
cardio.data <- cardio.data[cardio.data$ap_lo > 0,]

summary(cardio.data$ap_lo)
```


```{r}
glimpse(cardio.data)
```


# Research question 

My research question is: **Are gender, age, body weight, body height, blood pressure, cholesterol, glucose levels, smoking, drinking alcohol and activity level of an individual associated with the likelihood of contracting cardiovascular disease?**

I plan to use a logistic regression model to answer this research question since the target variable is a binary outcome.

# Cases 

The cases are the number of people who participate in the medical examination. There were a total of 70,000 cases in the original data file. After data pre-processing, the number of cases is 68,985. This change is due to the removal of rows that seemed to be errors such as extremely high and low blood pressure (-1,000 or 15,000). 

```{r}
nrow(cardio.data)
```


# Data collection 

The data was collected from medical information given by patient and examination results. "All of the dataset values were collected at the moment of medical examination." (https://www.kaggle.com/sulianova/cardiovascular-disease-dataset)

The data was downloaded from Kaggle (https://www.kaggle.com/sulianova/cardiovascular-disease-dataset) and then I uploaded it to Github to be used to import the data into R.

# Type of study 

This is an observational study since the analysis is on events that have already occurred. 

# Data Source 

The link to where I retrieved the data is: https://www.kaggle.com/sulianova/cardiovascular-disease-dataset

# Dependent Variable

The response variable is `cardio`. This is a qualitative variable since it is a categorical binary variable. `cardio` is an indicator variable that indicates whether or not someone has cardiovascular disease.

# Independent Variable

There are multiple variables that I am considering for analysis. The list contains a group of both quantitative and qualitative variables:

* `age` (quantitative): Age of patient in years
* `gender` (qualitative): Gender of patient
* `height` (quantitative): Height of patient in cm
* `weight` (quantitative): Weight of patient in kg
* `ap_hi` (quantitative): Systolic blood pressure
* `ap_lo` (quantitative): Diastolic blood pressure
* `cholesterol` (qualitative): Cholesterol level of patient
* `gluc` (qualitative): Glucose level of patient
* `smoke` (qualitative): Binary variable to determine if a patient smokes
* `alco` (qualitative): Binary variable to determine if a patient drinks alcohol
* `active` (qualitative): Yes/No if patient is physically active

# Relevant Summary Statistics 

## Summary Statistics

```{r}
summary(cardio.data)
```

--------------------------------------------------------------------------------

## Visualizations

This ggpairs plot of all the variables present in the dataset I wish to analyze is extremely difficult to read due to the number of variables. I will break the graph into smaller pieces below to further explore the data.

```{r}
entire.plot <- ggpairs(cardio.data)

entire.plot
```


### Cardiovascular Disease Outcome vs All Variables

#### Boxplots

* **Age**: Age may be strongly associated with determining cardiovascular disease due to the boxplots (A) for the two categories differing from one another.
* **Height**: Height may not be strongly associated with determining cardiovascular disease due to the boxplots (B) for the two categories appear to be almost identical. There are a large number of outliers for Height.
* **Weight**: Weight may be strongly associated with determining cardiovascular disease due to the boxplots (C) for the two categories differing from one another. There are a large number of outliers for Weight.
* **Systolic blood pressure**: Systolic blood pressure may be strongly associated with determining cardiovascular disease due to the boxplots (D) for the two categories differing from one another.
* **Diastolic blood pressure**: Diastolic blood pressure may be strongly associated with determining cardiovascular disease due to the boxplots (E) for the two categories differing from one another.

```{r}
library(ggpubr)

# Age
age1 <- ggplot(cardio.data, aes(x = cardio, y = age)) +
  geom_boxplot() +
  ggtitle("Boxplot: Age")

# Height
height1 <- ggplot(cardio.data, aes(x = cardio, y = height)) +
  geom_boxplot() +
  ggtitle("Boxplot: Height")

# Weight
weight1 <- ggplot(cardio.data, aes(x = cardio, y = weight)) +
  geom_boxplot() +
  ggtitle("Boxplot: Weight")

# Systolic blood pressure
sbp <- ggplot(cardio.data, aes(x = cardio, y = ap_hi)) +
  geom_boxplot() +
  ggtitle("Boxplot: Systolic blood pressure")

# Diastolic blood pressure
dbp <- ggplot(cardio.data, aes(x = cardio, y = ap_lo)) +
  geom_boxplot() +
  ggtitle("Boxplot: Diastolic blood pressure")

# Put plots together
ggarrange(age1, height1, weight1, sbp, dbp,
          labels = c("A","B","C","D","E"),
          ncol = 3, nrow = 2)
```

\clearpage

#### Histograms and Bar Plots

##### Age, Gender, Height, Weight

* **Age**: A larger portion of people who have cardiovascular disease are older than people who do not have cardiovascular disease, according to the histograms below.
* **Gender**; There appears to be a substantially larger number of females than males present in this dataset. I wonder if this implies that women are more prone to heart conditions. The number of Male and Female partipants seems to be fairly equal in both factor levels of `cardio`
* **Height**: For both factor levels of `cardio`, the `height` appear to follow a bell curve
* **Weight**: For both factor levels of `cardio`, most of the weights fall between 50 and 100 kg. 

```{r}
primary_var <- "cardio"
pairs <- ggpairs(cardio.data, columns = c(1:4,12))
pvar_pos <- match(primary_var, pairs$xAxisLabels)
plots <- lapply(1:pairs$ncol, function(j) getPlot(pairs, i = pvar_pos, j = j))
matrix1 <- ggmatrix(
    plots,
    nrow = 1,
    ncol = pairs$ncol-1,
    xAxisLabels = pairs$xAxisLabels[1:4],
    title = "Cardiovascular Disease vs Age, Gender, Height, and Weight"
)


matrix1
```


##### Systolic blood pressure, Diastolic blood pressure, Cholesterol, Smoke

* **Systolic blood pressure**: A large portion of the data 100 and 150. For people who don't have cardiovascular disease, there is a unimodal peak that is about 2 to 2.5 times the length of any other histogram bin.
* **Diastolic blood pressure**: A large portion of the data 50 and 100. For people who don't have cardiovascular disease, there is a unimodal peak that is about 2 to 2.5 times the length of any other histogram bin.
* **Cholesterol**: There is an unequal distribution of factor levels of cholesterol. Most patients have a normal cholesterol level.
* **Glucose**: There is an unequal distribution of factor levels of glucose. Most patients have a normal glucose level.

```{r}
primary_var <- "cardio"
pairs <- ggpairs(cardio.data, columns = c(5:8,12))
pvar_pos <- match(primary_var, pairs$xAxisLabels)
plots <- lapply(1:pairs$ncol, function(j) getPlot(pairs, i = pvar_pos, j = j))
matrix1 <- ggmatrix(
    plots,
    nrow = 1,
    ncol = pairs$ncol-1,
    xAxisLabels = pairs$xAxisLabels[1:4],
    title = "Cardiovascular Disease vs Systolic blood pressure, 
    Diastolic blood pressure, Cholesterol, Glucose"
)


matrix1
```


##### Smoke, Alcohol, and Activity level:

For all 3 variables, there is an uneven distribution of factor levels. There is a much larger proportion of non smokers in this dataset. The dataset contains mostly non alcoholics. The dataset has a larger number of active people.

```{r}
primary_var <- "cardio"
pairs <- ggpairs(cardio.data, columns = c(9:11,12))
pvar_pos <- match(primary_var, pairs$xAxisLabels)
plots <- lapply(1:pairs$ncol, function(j) getPlot(pairs, i = pvar_pos, j = j))
matrix1 <- ggmatrix(
    plots,
    nrow = 1,
    ncol = pairs$ncol-1,
    xAxisLabels = pairs$xAxisLabels[1:3],
    title = "Cardiovascular Disease vs Smoke, 
    Alcohol, Activity Level"
)


matrix1
```

\clearpage

### Numerical Variables: Scatterplots, Density Plots, Correlation Matrix

There does not appear to be any strong correlation between any of the numeric variables I plan to use as independent variables. The strongest correlation exists between the Systolic blood pressure and Diastolic blood pressure with a correlation of 0.65. 

Since the explanatory variables are not strongly correlated with each other, there is not an issue with multicollinearity in the data/model.

```{r}
ggpairs(cardio.data, columns = c(1,3:6))
```

\clearpage

### Mosaic Plots

#### Cardiovascular Disease Outcome by Age

```{r}
plot(cardio.data$cardio ~ cardio.data$age,
     xlab = "Age",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Age")
```


--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Gender

```{r}
plot(cardio.data$cardio ~ cardio.data$gender,
     xlab = "Gender",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Gender")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Height

```{r}
plot(cardio.data$cardio ~ cardio.data$height,
     xlab = "Height",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Height")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Weight

```{r}
plot(cardio.data$cardio ~ cardio.data$weight,
     xlab = "Weight",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Weight")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Systolic blood pressure

```{r}
plot(cardio.data$cardio ~ cardio.data$ap_hi,
     xlab = "Systolic Blood Pressure",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Systolic Blood Pressure")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Diastolic blood pressure

```{r}
plot(cardio.data$cardio ~ cardio.data$ap_hi,
     xlab = "Diastolic Blood Pressure",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Diastolic Blood Pressure")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Cholesterol Level

```{r}
plot(cardio.data$cardio ~ cardio.data$cholesterol,
     xlab = "Cholesterol Level",ylab = "Cardio", 
     main = "Cardiovascular Outcome vs. Cholesterol Level")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Glucose Level

```{r}
plot(cardio.data$cardio ~ cardio.data$gluc,
     xlab = "Glucose Level",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Glucose Level")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Smoking Level

```{r}
plot(cardio.data$cardio ~ cardio.data$smoke, 
     xlab = "Smoking",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Smoking")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Alcohol Level

```{r}
plot(cardio.data$cardio ~ cardio.data$alco,
     xlab = "Alcohol",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Alcohol Level")
```

--------------------------------------------------------------------------------

\clearpage

#### Cardiovascular Disease Outcome by Activity Level

```{r}
plot(cardio.data$cardio ~ cardio.data$active,
     xlab = "Active",ylab = "Cardio",
     main = "Cardiovascular Outcome vs. Activity Level")
```

--------------------------------------------------------------------------------